---
title: "Investigating the effects of race and gender"
author: "Odell, Christopher"
date: '2020-02-02'
output:
  word_document: default
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(dplyr)
library(ggplot2)
library(tree)
#install.packages("Epi")
#library(parcor)
#library(Epi)

data <- read.table("psam_p41.csv", header=T, quote="\"", sep=",")

q_data<- data %>% dplyr::select("SEX", "AGEP", "SCHL", "RAC1P", "MAR", "WKHP", "PINCP", "OCCP", "COW")
q_data <- as_tibble(q_data)
summary(q_data)
q_data_fitlered <- q_data %>% filter(AGEP > 16 , AGEP < 67, !is.na(SCHL),!is.na(WKHP), PINCP >0)
summary(q_data %>% filter(AGEP > 16 , AGEP < 67))
q_data_fitlered$RAC1P <- as.factor(q_data_fitlered$RAC1P)
q_data_fitlered$SEX <- as.factor(q_data_fitlered$SEX)
q_data_fitlered$MAR <- as.factor(q_data_fitlered$MAR)
# q_data$OCCP <- as.factor(q_data$OCCP)
q_data_fitlered$COW <- as.factor(q_data_fitlered$COW)
q_data_fitlered$OCCP
q_data_fitlered$OCCP <- cut(q_data_fitlered$OCCP, 
                            breaks=c(0, 430, 740, 950, 1240,1560,
                                     1965,2060,2160,2550,2920,3540,
                                     3655,3955,4150,4250,4650, 4965,
                                     5940,6130,6765,6940,7630,8965,
                                     9750,9830,Inf),
                            labels=c("MGR", "BUS","FIN","CMM","ENG","SCI",
                                         "CMS", "LGL", "EDU", "ENT", "MED", "HLS",
                                         "PRT","EAT", "CLN", "PRS","SAL", "OFF", 
                                         "FFF","CON", "EXT", "RPR", "PRD","TRN",
                                         "MIL","OTHER"))
#class(q_data_fitlered$OCCP)
ethnicity <- c("White alone", "African American alone", "American Indian alone", "Alaska Native alone", "Native American", "Asian alone", "Native Hawaiian alone", "Other", "Two or More Races")

levels(q_data_fitlered$RAC1P) <- ethnicity
levels(q_data_fitlered$SEX) <- c("Male", "Female")

```


# To find potential relationship between varaibles and proper forms of variables included in linear models, based on following plots for Q1-3:

PINCP VS RAC1P
```{r}
#ggplot(q_data_fitlered,aes(RAC1P,PINCP))+geom_boxplot()
ggplot(q_data_fitlered,aes(RAC1P,log(PINCP)))+ 
  theme(axis.text.x = element_text(angle = 90)) +geom_boxplot()
```
Minor differences are observed, log form for PINCH is more desired.

PINCP VS SEX
```{r}

#ggplot(q_data_fitlered,aes(SEX,PINCP))+geom_boxplot()
ggplot(q_data_fitlered,aes(SEX,log(PINCP)))+geom_boxplot()

```
Slight differnces could be observed by sex. Again log form for PINCH is more desired.

PINCP VS WKHP
```{r}
#qplot(WKHP, PINCP, data = q_data_fitlered)
qplot(WKHP, log(PINCP), data = q_data_fitlered)
```
Linearity could be observed: WKHP VS LOG(PINCP)

PINCP VS SCHL
```{r}
#qplot(SCHL, PINCP, data = q_data_fitlered)
#qplot(SCHL^2, PINCP, data = q_data_fitlered)
#ggplot(q_data_fitlered,aes(SCHL,log(PINCP)))+ geom_point()+
 # stat_summary(aes(y=log(PINCP),group=1), fun.y=mean,geom='line',color="red",group=1)
ggplot(q_data_fitlered,aes(SCHL^2,log(PINCP)))+ geom_point()+
  stat_summary(aes(y=log(PINCP),group=1), fun.y=mean,geom='line',color="red",group=1)
q_data_fitlered$SCHL2 <- (q_data_fitlered$SCHL)^2 
```
SCHL in quadratic form shows linearity with log(PINCP).

PINCP VS OCCP
```{r}

ggplot(q_data_fitlered,aes(x=reorder(q_data_fitlered$OCCP, q_data_fitlered$PINCP, mean),log(PINCP), fill=SEX))+
geom_boxplot()+ 
scale_fill_manual(labels=c("Male", "Female"),values=c("green", "light blue"))+
theme(axis.text.x = element_text(angle = 45, hjust = 1),legend.position = "bottom" )+
labs(title = "Log Transformed Personal Income by Occupation", subtitle = "Subset by Gender", y ="Log Transformed Personal Income", x="Grouped Occupation")
```
Differences could be observed.

PINCP VS COW

```{r}
ggplot(q_data_fitlered,aes(COW,log(PINCP)))+geom_boxplot()
```
There are differences between different categories.

PINCP VS MAR

```{r}
ggplot(q_data_fitlered,aes(MAR,log(PINCP)))+geom_boxplot()
```
Slight differences could be observed.

PINCP VS AGE
```{r}
qplot(AGEP, log(PINCP), data = q_data_fitlered)
```
Roughly linear relationship could be observed.

SCHL versus WKHP

```{r}

q4_plot_data <- q_data_fitlered

#Verify that it filtered
print(q4_plot_data)

#Setting the boundaries for the categories
max(q4_plot_data$WKHP)
min(q4_plot_data$WKHP)
max(q4_plot_data$PINCP)
min(q4_plot_data$PINCP)

#q4_plot_data$WKHP<-cut(q4_plot_data$WKHP, c(0, 10, 20, 30, 40, 50, 60, 70, 99))
q4_plot_data$PINCP<-cut(q4_plot_data$PINCP, c(-15000, 25000, 50000, 75000, 10000, 150000, 200000, 300000, 1500000))

#print(q4_plot_data)

#q4_plot_data %>% 
#  ggplot()+
#  geom_bar(aes(WKHP ,..count..))

q4_plot_data %>% 
  ggplot()+
  geom_boxplot(aes(PINCP ,WKHP))+
  scale_x_discrete(labels=c("(-15K to 25K]", "(25K to 50K]", "(50K to 75K]", "(75K to 100K]", "(100K to 150K]", "(150K to 200K]", "(200K to 300K]", "(300K to 1,500K]")) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),plot.title = element_text(hjust = 0.5))+
  labs(x="Personal Income", y="Hours of Work Per Week",title="Distribution of Hours Worked by Person Income Categories",
       caption="The data is based on individuals living in Oregon, with the age range of 16 to 67")
```


# Model Fitting

Model Fitting

Setting up the data, filtering for consistency and taking out 25% for test and inference

```{r}
set.seed(2020)

(n <- dim(q_data_fitlered)[1])
(r <- round(n * .75))
idx <- 1:n
nidx <- sample(idx, r, replace = FALSE)
data_train <- q_data_fitlered[nidx, ]
data_test <- q_data_fitlered[-nidx, ]

```

```{r}

final_q1_q3 <- lm(formula = log(PINCP) ~ WKHP + OCCP + AGEP + SCHL2 + COW + 
    MAR + SEX + SCHL + RAC1P, data = q_data_fitlered)

final_q4_a <- lm(formula = WKHP ~ log(PINCP) + OCCP + SEX + AGEP + COW + MAR + RAC1P + SCHL,
               data = data_train)

summary(final_q1_q3)
summary(final_q4_a)

```




## Question 1 
```{r}
q1_lower <- lm(log(PINCP) ~ 1, data=data_train)
q1_upper <- lm(log(PINCP) ~ SEX + AGEP + SCHL + SCHL2 + MAR + WKHP + COW + OCCP, data=data_train)
step(q1_lower, scope=list(upper=q1_upper, lower=q1_lower),direction="forward",test="F")

q1_tree <- data_train

q1_tree$PINCP2<-cut(q1_tree$PINCP, c(0, 25000, 50000, 75000, 10000, 150000, 200000, 300000, 1500000))
tr <- tree(PINCP2 ~ COW  + WKHP + SCHL + AGEP + SEX + MAR + RAC1P, data = q1_tree)
summary(tr)

plot(tr); text(tr)



```

```{R}
# Final model
# + RAC1P
final_q1 <- lm(formula = log(PINCP) ~ WKHP + OCCP + AGEP + SCHL2 + COW + 
    MAR + SEX + SCHL + RAC1P, data = data_train)

# Residual vs Fitted
qplot(final_q1 $fitted.values, final_q1 $residuals)

# Residual VS Explanatory
qplot(data_train$WKHP, final_q1$residuals)
qplot(data_train$AGEP, final_q1$residuals)
qplot(data_train$SCHL, final_q1$residuals)

# QQ Plot
par(mfrow = c(1, 1))
qqnorm(final_q1$residuals)
qqline(final_q1$residuals)
```
The plot of Residual vs Fitted looks good, but normality is violated.

```{R}
# To answer Q1
ans_q1 <- lm(formula = log(PINCP) ~ WKHP + OCCP + AGEP + SCHL2 + COW + 
    MAR + SEX + SCHL + RAC1P, data = data_test)
summary(ans_q1)
# Residual vs Fitted
qplot(ans_q1$fitted.values, ans_q1$residuals)

# Residual VS Explanatory
qplot(data_test$WKHP, ans_q1$residuals)
qplot(data_test$AGEP, ans_q1$residuals)
qplot(data_test$SCHL, ans_q1$residuals)
qplot(data_test$RAC1P, ans_q1$residuals)


# QQ Plot
par(mfrow = c(1, 1))
qqnorm(ans_q1$residuals)
qqline(ans_q1$residuals)
```

# Question 2
```{r}
q2_lower <- lm(log(PINCP) ~ 1, data=data_train)
q2_upper <- lm(log(PINCP) ~ RAC1P + AGEP + SCHL + SCHL2 + MAR + WKHP + COW + OCCP, data=data_train)
step(q2_lower, scope=list(upper=q2_upper, lower=q2_lower),direction="forward", test="F")
```

```{R}
# Final model
# + SEX
final_q2 <- lm(formula = log(PINCP) ~ WKHP + OCCP + AGEP + SCHL2 + COW + 
    MAR + SCHL + RAC1P + SEX, data = data_train)


test <- lm(formula = log(PINCP) ~WKHP + OCCP, data = data_train)
anova(test, final_q2)

# Residual vs Fitted
qplot(final_q2$fitted.values, final_q2$residuals)

# Residual VS Explanatory
qplot(data_train$WKHP, final_q2$residuals)
qplot(data_train$AGEP, final_q2$residuals)
qplot(data_train$SCHL, final_q2$residuals)
qplot(data_train$SCHL2, final_q2$residuals)
# QQ Plot
par(mfrow = c(1, 1))
qqnorm(final_q2$residuals)
qqline(final_q2$residuals)
```
Fitted resisual plot looks good, normality is violated.

```{R}
# To answer Q2
ans_q2 <- lm(formula = log(PINCP) ~ WKHP + OCCP + AGEP + SCHL2 + COW + 
    MAR + SCHL + RAC1P + SEX, data = data_test)
summary(ans_q2)
# Residual vs Fitted
qplot(ans_q2$fitted.values, ans_q2$residuals)

# Residual VS Explanatory
qplot(data_test$WKHP, ans_q2$residuals)
qplot(data_test$AGEP, ans_q2$residuals)
qplot(data_test$SCHL, ans_q2$residuals)
qplot(data_test$SEX, ans_q2$residuals)


# QQ Plot
par(mfrow = c(1, 1))
qqnorm(ans_q2$residuals)
qqline(ans_q2$residuals)
```

# Question 3
```{r}
q3_lower <- lm(log(PINCP) ~ 1, data=data_train)
q3_upper <- lm(log(PINCP) ~ SEX + AGEP + MAR + COW + OCCP + RAC1P, data=data_train)
step(q3_lower, scope=list(upper=q3_upper, lower=q3_lower),direction="forward", test="F")

```

```{R}
# Final model
# + SCHL, +WKHP
final_q3 <- lm(formula = log(PINCP) ~ OCCP + AGEP + COW + MAR + SEX + RAC1P + SCHL+ SCHL2 + WKHP, data = data_train)

# Residual vs Fitted
qplot(final_q3$fitted.values, final_q3$residuals)

# Residual VS Explanatory
qplot(data_train$WKHP, final_q3$residuals)
qplot(data_train$AGEP, final_q3$residuals)
qplot(data_train$SCHL, final_q3$residuals)
qplot(data_train$SCHL2, final_q3$residuals)
# QQ Plot
par(mfrow = c(1, 1))
qqnorm(final_q3$residuals)
qqline(final_q3$residuals)
```
Again, fitted resisual plot looks good, normality is violated.

```{R}
# To answer Q3
ans_q3 <- lm(formula = log(PINCP) ~ OCCP + AGEP + COW + MAR + SEX + RAC1P + SCHL + SCHL2 + WKHP, data = data_test)

test <- lm(formula = log(PINCP) ~ OCCP + AGEP + COW + MAR + SEX + RAC1P, data = data_test)

summary(ans_q3)

anova(test, ans_q3)

# Residual vs Fitted
qplot(ans_q3$fitted.values, ans_q3$residuals)

# Residual VS Explanatory
qplot(data_test$AGEP, ans_q3$residuals)
qplot(data_test$SCHL, ans_q3$residuals)
qplot(data_test$WKHP, ans_q3$residuals)

# QQ Plot
par(mfrow = c(1, 1))
qqnorm(ans_q3$residuals)
qqline(ans_q3$residuals)
```

Issues for 1-3: 
Normality is violated;
Constant vairance may be violated;
All three questions are answered by the full model, train set is too large?

# Question 4

# To find potential relationship between varaibles and proper forms of variables included in linear models, based on following plots for Q4:

```{r}
ggplot(q_data_fitlered,aes(SEX,WKHP))+geom_boxplot()

ggplot(q_data_fitlered,aes(OCCP,WKHP))+geom_boxplot()+ theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggplot(q_data_fitlered,aes(COW,WKHP))+geom_boxplot()

ggplot(q_data_fitlered,aes(RAC1P,WKHP))+geom_boxplot()

ggplot(q_data_fitlered,aes(MAR,WKHP))+geom_boxplot()

ggplot(q_data_fitlered,aes(AGEP, WKHP))+ geom_point()+
  stat_summary(aes(y=log(PINCP),group=1), fun.y=mean,geom='line',color="red",group=1)

ggplot(q_data_fitlered,aes(SCHL, WKHP))+ geom_point()+
  stat_summary(aes(y=SCHL,group=1), fun.y=mean,geom='line',color="red",group=1)

ggplot(q_data_fitlered,aes(log(PINCP), WKHP))+ geom_point()+
  stat_summary(aes(y=log(PINCP),group=1), fun.y=mean,geom='line',color="red",group=1)

```

# Question 4
```{r}
q4_lower <- lm(WKHP ~ 1, data=data_train)
q4_upper <- lm(WKHP ~ SEX + AGEP + MAR + COW + log(PINCP) + RAC1P + OCCP, data=data_train)
step(q4_lower, scope=list(upper=q4_upper, lower=q4_lower),direction="forward")

```

```{R}
# Final model
# + SCHL
final_q4 <- lm(formula = WKHP ~ log(PINCP) + OCCP + SEX + AGEP + COW + MAR + RAC1P + SCHL,
               data = data_train)
test <- lm(formula = WKHP ~ log(PINCP) + OCCP + SEX + AGEP + COW + MAR +  RAC1P + SCHL,
               data = data_train)
summary(final_q4)
anova(test, final_q4)

# Residual vs Fitted
qplot(final_q4$fitted.values, final_q4$residuals)

# Residual VS Explanatory
log_PINCP <- log(data_train$PINCP)
qplot(log_PINCP, final_q4$residuals)
qplot(data_train$AGEP, final_q4$residuals)
qplot(data_train$SCHL, final_q4$residuals)

# QQ Plot
par(mfrow = c(1, 1))
qqnorm(final_q4$residuals)
qqline(final_q4$residuals)
```
patterns of residual is not explained by current variables.

```{R}
# To answer Q4
ans_q4 <- lm(formula = WKHP ~ log(PINCP) + OCCP + SEX + AGEP + COW + MAR + RAC1P + SCHL, data = data_test)
summary(ans_q4)
# Residual vs Fitted
qplot(ans_q4$fitted.values, ans_q4$residuals)

# Residual VS Explanatory
qplot(data_test$WKHP, ans_q4$residuals)
qplot(data_test$AGEP, ans_q4$residuals)
qplot(data_test$SCHL, ans_q4$residuals)

# QQ Plot
par(mfrow = c(1, 1))
qqnorm(ans_q4$residuals)
qqline(ans_q4$residuals)
```

```{R}
# Try penalized models

#install.packages("plotmo")
library(glmnet)
library(ncvreg)
library(parcor)
library(mht)
library(plotmo)

# Q1-Q3
x <- data.matrix(data_train[,1:6])
z <- data.matrix(data_train[,8:9])
x <- cbind(x, z)
y <- log(as.matrix(data_train[,7]))

## issue: factors are suspected to treat as numerical???
elnet <- glmnet(x, y, alpha = 0.5)
elnet.cv <- cv.glmnet(x, y, alpha = 0.5)
# Finding the Elastic net estimates for lambda with minimum cross-validation error
elnet1 <- glmnet(x, y, lambda = elnet.cv$lambda.min, alpha = 0.5)
# Elastic net estimates of coefficients
elnet1$beta

plotres(elnet.cv)

##### Adaptive Lasso Estimates
adlasso <- adalasso(x, y)
adlasso$coefficients.adalasso
adlasso


# This part is not quite right, trying to plot the residuals
#x_test <- data.matrix(data_test[,1:6])
#z_test <- data.matrix(data_test[,8:9])
#x_test <- cbind(x_test, z_test)

#preds <- predict(elnet1, newx=x_test)
#test1 <- data_test
#test1
#test1$preds <- preds
#test1$residuals <- test1$preds-test1$PINCP

#qplot(log(test1$PINCP), test1$residuals)


# Q4
x_q4 <- data.matrix(data_train[,1:5])
z_q4 <- data.matrix(data_train[,7:9])
x_q4 <- cbind(x_q4, z_q4)
y_q4 <- as.matrix(data_train[,6])

## issue: factors are suspected to treat as numerical???
elnet_q4 <- glmnet(x_q4, y_q4, alpha = 0.5)
elnet.cv_q4 <- cv.glmnet(x_q4, y_q4, alpha = 0.5)
# Finding the Elastic net estimates for lambda with minimum cross-validation error
elnet1_q4 <- glmnet(x_q4, y_q4, lambda = elnet.cv$lambda.min, alpha = 0.5)
# Elastic net estimates of coefficients
elnet1_q4$beta

plotres(elnet.cv_q4)

##### Adaptive Lasso Estimates
adlasso_q4 <- adalasso(x, y)
adlasso_q4$coefficients.adalasso
adlasso_q4
```


#Appendix A - Source Code
```{r code=readLines(knitr::purl('./Project1_report_Code.Rmd', documentation = 0)), eval = FALSE, echo=TRUE}

```